Discussion

### 2020-07-27 text/discussion.Rmd

One of the biggest hurdles in cancer research is the size of the search space;
~20,000 protein-coding genes is comparable with the number of tumor samples,
even with multiple mutations per sample. And that is not accounting for the fact
that ~30k non-coding genes and hundreds of thousands of non-coding regulatory
elements exist with oncogenic or tumor suppressor activity [@deesMuSiC2012;
@tamboreroOncodriveCLUST2013;
@lawrenceDiscoverySaturationAnalysis2014;
@kumarStatisticallyIdentifyingTumor2015; @jiangWITER2019;
@zhaoModelingPositive2019; @non-coding-mutations-Zhang]. Hence, we sought to
simplify the problem by investigating cancer as a disease of basic cellular and
biochemical pathways, which we accomplished by translating gene-level mutations
into pathway level disruptions. Our approach differs from previously proposed
methods [@PATHNET] in that we chose to focus on the pathway as the unit of
disruption instead of the gene, employing the conceit that individual mutations
may be sufficient to disrupt entire pathways (as is often the case with
drivers). Rather than evaluating the burden of mutations on each pathway within
or across tumors, which can be biased by the number of genes or the coverage of
pathway-linked genes in the genome, and which we find to be relatively uniformly
distributed, we scored each pathway in binary fashion, allowing one mutation or
multiple mutations to count equally for each sample. To our knowledge, this
approach has not previously been attempted in the literature in spite of its
relative simplicity. We limited our analysis to mutations with likely
deleterious effects, including missense, nonsense, frame-shift, stop-gain and
splice site mutations in genes that are actively expressed in each cancer type,
thus avoiding bias from transcription coupled repair. Our method of filtration
differs from the "rank-and-cut" method of [@PCAWG-MAIN] but represents a
reasonable attempt to account for the same biases. We also filtered genes on the
basis of pathway membership in biochemical pathways, excluding curated genesets
related to diseases, syndromes, or classes of proteins ( _e.g._ transcription
factors). We chose this approach to limit redundancy, which is already high in
the remaining pathways for some driver genes, and to exclude biologically
unrelated collections of genes. We do not account for non-coding mutations,
amplification events or structural variants, though we hope to include them in
subsequent studies (see below). Our pathway-centric analysis also allows us to
circumvent the problem that some cancer-promoting mutations likely occur at low
enough frequencies as to be indistinguishable from background. Previous attempts
to capture these associations better using Bayesian networks have yielded
important new insights into this problem and new genes
[@hornNetSigNetworkbasedDiscovery2018] that would not have been possible with
traditional analysis, supporting the case that pathway-centric approaches like
ours have the potential to increase statistical power.

The hypothesis that cancer is the result of dysfunction from
a limited number of basic cellular processes common to all living human cells
was first introduced and later expanded on in a pair of classic essays by
Hanahan and Weinberg [@weinberg200; @hallmarks_II]. In this view, all tumors
regardless of tissue of origin acquire six essential phenotypes; sustaining
proliferative signaling, evading growth suppressors, replicative immortality,
evading apoptosis, supportive angiogenesis, tissue invasion and metastasis
[@weinberg2000]. Advancements in the field over the ensuing decade led to the
inclusion of new tumor promoting processes, namely deregulation of cellular
energy cycle, tumor promoting inflammation, evasion of the immune system, genome
instability, and on top of these acquired traits a new appreciation for the
concept of the tumor microenvironment and its contribution to the evolution and
promotion of metastatic tumorigenesis [@hallmarks_II].

@@

We find that most tumors segregate into one of twelve biologically distinct
clusters, regardless of the cancer type (**figure 1G,I** and **supplementary
figure S1**). Some cancer types ( _e.g._ pancreatic adenocarcinomas in the RAS
mutant clusters 8 and 9, and Kidney Chromophobe in cluster 1, see **supplementary
figure S1**) are unevenly distributed, though we could not identify any cancers
that were exclusive to a single tumor class. Only cancer types with the fewest
samples were found to be absent from one or more classes at all. Somewhat
surprisingly to us, this finding extends to histological subtypes of breast,
head and neck cancers, leukemias, _etc._ (see **Figure 1J** and **Supplemental
Figure S2**). This result implies that many, if not most, histological subtypes
could reflect differences in cell-of-origin, rather than fundamental differences
in proximal cause. The four major subtypes of breast cancer correspond to histological
and molecular expression profiles that define them and how they respond to experimental
stimuli [@breast-subtypes2]. It has been hypothesized that differences in the
molecular regulators of development in the precursor cell types present in the
breast epithelium drive the various histological phenotypes [@breast-subtypes;
@breast-subtypes3]. Consistent with this view we found that breast tumor samples
of the Luminal A subtype were heavily biased toward membership in clusters 1 and
7, and basal tumor samples were biased toward cluster 6 (**figure 1J**).
However, both subtypes also contained samples in every other class (without
exception), and Luminal B and Her2 positive samples are distributed across
clusters. We interpret these data to mean that inherited cell-of-origin
molecular profiles may predispose certain precursor cell types within the breast
epithelium to forming tumors of one class or another but is not determinative;
this view is compatible with the previously stated hypothesis but opens the way
for a more granular view of individual tumors. It would be surprising if we did
not observe bias for some cancers and subtypes amongst our classes, since some
treatment regimens have greater efficacy for patients
of a given cancer or histological subtype [@breast-subtypes2]. Nonetheless, the
basis for some tumors being treatment-refractory in spite of receiving the
standard of clinical care for diagnostic markers remains elusive. Doubtless some
of this is due to bad luck, as tumors can metastasize and remain dormant years
before they are detected at distal sites, or resistant clones may have already
arisen at undetectable levels [@hallmarks_II], but our analysis suggests the
possibility of identifying more informative molecular, histological or cellular
subtypes that could form a basis for future separation of patients into
different precision treatment regimens.

@@

Finally, we must remark on the limitations of our work exemplified in two of the
classes, 1 and 10, which for which we did not find distinctive associations with
pathways (**figure 2B, C**). Class 1 has a relatively low proportion of mutated
pathways across the board, although it is broadly enriched in many of the same
tumor-promoting pathways common to the other groups. This likely does not
reflect a true difference in stages of tumor evolution, as our data show clearly
that this class is as likely to contain stage IV metastatic tumors as it is to
contain those of stage I (**figure 1L, M**). This class likely represents a
group of tumors with aberrations in methylation, copy number alterations,
structural variants, or combinations thereof which we excluded from our
analysis. Consistent with this, kidney chromophobe and thyroid cancers have high
proportions of SVs _vs._ other variant types [@PCAWG-MAIN] and are heavily
skewed to cluster 1 membership (**Supplementary figure S1**). Likewise, Class 8
represents a group of hyper-mutated tumors that harbor so many mutations that
virtually no pathway is unaffected. It seems likely that a significant fraction
of the samples flagged as “mutant” for each pathway are actually unaffected,
consistent with those samples have low information content. This could be
addressed with application of ever-increasingly sophisticated bioinformatics
algorithms for filtering likely passenger mutations [ _e.g._ @spliceAI;
@deep-learning-clinical-variants]. In the future, we hope to incorporate these
other data into a comprehensive pathway-centric analysis as we have done here
for point mutations and indels.

@@

Our findings imply that there are separate processes in the etiology of cancer
that can be broadly thought of as general cancer promoting, class-specific
mutations and metastasis. General cancer promoting processes must include
factors that relate to genome stability and immortality, as "enabling
characteristics" of the cancer phenotype [@hallmarks_II]. Such pathways are
disrupted in all or most of the defined classes (**supplementary figure S1**)
and often are the result of aberrations involving common driver genes such as
BRCA1/2, MMR genes, mitotic checkpoints, Cohesion complexes, _etc._
(**supplementary figure S3** and **supplemental file 1**). Class-specific
evolution must involve the acquisition of disruptions to pathways that may
individually be harmful ( _e.g._ highly proliferative cells are more likely to
senesce) but together produce more specialized cancer phenotype and increased
fitness (as in **figure 4** and **tables 1** and **ST2**). Importantly, our work
does not distinguish between the order in which these mutations should
accumulate. This could be addressed in a future study by integrating our
analysis with evolutionary analysis of clonality, drawing inferences from
variant allele frequencies as in @PCAWG-EVO. However, since many of the genes in
the non-cluster-specific pathways involve genome stability, it is reasonable to
hypothesize that these mutations promote or enable the acquisition of
class-specific defects via random mutation and natural selection, thus producing
the classes we observed. In support of this, the pan-cancer analysis of whole
genomes consortium (PCAWG) found that a subset of known oncogenic driver
mutations are highly enriched in early arising clones, whereas later arising
clones have much greater diversity in driver mutations [@PCAWG-EVO]. Moreover,
driver genes that are known to be responsible for discrete mutation signatures
such as APOBEC, BRCA1 and BRCA2 produce mutational hotspots reflecting the
varying selective pressures in different tissues [@PCAWG-MAIN].

One of the drawbacks of bulk tumor whole genome sequencing data is the problem
of tumor heterogeneity. Consortium samples are likely to contain contamination
from support tissue and inflammatory and immune cells of the innate and adaptive
immune systems, potentially harboring cancer supporting mutations
[@bhowmick_review]. We think it will be instructive to explore these ideas in
the context of single cell experiments.

@@

We report that pathway enrichment in metastatic tumors across all classes
(pooled) yielded generally lower effect sizes and larger p-values than the
cluster specific analysis, suggesting that signal is diluted when the samples
are analyzed this way, and supporting the view that metastasis has
class-specific requirements. Since the number of metastatic tumors is relatively
low this part of our analysis is likely underpowered and subject to expanded
analysis with larger cohorts. Nonetheless, the class-specific metastatic pathway
enrichment we observe (in **table 2**) could be a result of the pathways
directly promoting metastatic phenotypes in these classes, or because they
acquire a stronger or more complete oncogenic phenotype, as we don’t know the
order in which they were acquired in individual tumor samples. The fact that
most metastatic enrichment is cluster-specific and has a tendency to overlap
with cluster-specific pathways from non-metastatic tumors of neighboring
clusters favors the latter interpretation, and suggests a mechanism of
complementarity, in which newly acquired mutations result in similar classes
converging on a single deadly phenotype that captures most or all of the
critical features of end-stage cancer. One explanation for this could be that
tumors must accumulate mutations in a certain order or face elimination due to
any of the selective bottlenecks they face (senescence, the immune system,
_etc._) [@hallmarks_II]. Some clusters were characterized by cluster-specific
pathways that seem superficially unrelated; for example cluster 3 has mutations
impacting protein translation and turnover, whereas cluster 5 has defects in RNA
metabolism and clusters 7, 8 and 9 have defects in the more traditional
extra-cellular signaling pathways (as exemplified in **table 1**). It may be
that the primary effects of mutations in protein and RNA metabolism in clusters
3 and 5 point to currently unappreciated mechanisms impacting hallmark pathways.
Clusters 3 and 5 are also marked by a decidedly different set of pathways in
metastasis. Clusters 3 and 5 notably have (in addition to signaling pathways)
mutations in glucagon and glycogen metabolism, respectively, in their metastases
(**table 2**), consistent with energy metabolism as one of the new hallmarks
[@hallmarks_II], but also distinguishing these classes from the others in the
order or nature of hallmark acquisition. Thus, even at relatively low power, our
analysis of metastasis further supports pathway-centric analysis of tumors as
indicative of real differences in underlying biology. As a major caveat to this,
we acknowledge that many therapies _do_ target specific signaling pathways (
_e.g._ EGFR or HER2), but understanding the broader context of the genetic
background and pathway vulnerability of tumors containing such markers may aid
in creating smarter combination therapies. We submit that when we discover the
requirements of each cluster with respect to pathway disruptions and metastasis
we may be able to target them therapeutically and prevent further adaptation.

>>>>

One of the biggest hurdles in cancer research is the size of the search space;
~20,000 protein-coding genes is comparable with the number of tumor samples,
even with multiple mutations per sample. And that is not accounting for the fact
that ~30k non-coding genes and hundreds of thousands of non-coding regulatory
elements exist with oncogenic or tumor suppressor activity [@deesMuSiC2012;
@tamboreroOncodriveCLUST2013;
@lawrenceDiscoverySaturationAnalysis2014;
@kumarStatisticallyIdentifyingTumor2015; @jiangWITER2019;
@zhaoModelingPositive2019; @non-coding-mutations-Zhang]. Hence, we sought to
simplify the problem by investigating cancer as a disease of basic cellular and
biochemical pathways, which we accomplished by translating gene-level mutations
into pathway level disruptions. Our approach differs from previously proposed
methods [@PATHNET] in that we chose to focus on the pathway as the unit of
disruption instead of the gene, employing the conceit that individual mutations
may be sufficient to disrupt entire pathways (as is often the case with
drivers). Rather than evaluating the burden of mutations on each pathway within
or across tumors, which can be biased by the number of genes or the coverage of
pathway-linked genes in the genome, and which we find to be relatively uniformly
distributed, we scored each pathway in binary fashion, allowing one mutation or
multiple mutations to count equally for each sample.

To our knowledge, this approach has not previously been attempted in the
literature in spite of its relative simplicity. We limited our analysis to
mutations with likely deleterious effects, including missense, nonsense,
frame-shift, stop-gain and splice site mutations in genes that are actively
expressed in each cancer type, thus avoiding bias from transcription coupled
repair. Our method of filtration differs from the "rank-and-cut" method of
[@PCAWG-MAIN] but represents a reasonable attempt to account for the same
biases. We also filtered genes on the basis of pathway membership in biochemical
pathways, excluding curated genesets related to diseases, syndromes, or classes
of proteins ( _e.g._ transcription factors). We chose this approach to limit
redundancy, which is already high in the remaining pathways for some driver
genes, and to exclude biologically unrelated collections of genes. We do not
account for non-coding mutations, amplification events or structural variants,
though we hope to include them in subsequent studies (see below). Our
pathway-centric analysis also allows us to circumvent the problem that some
cancer-promoting mutations likely occur at low enough frequencies as to be
indistinguishable from background. Previous attempts to capture these
associations better using Bayesian networks have yielded important new insights
into this problem and new genes [@hornNetSigNetworkbasedDiscovery2018] that
would not have been possible with traditional analysis, supporting the case that
pathway-centric approaches like ours have the potential to increase statistical
power.

The hypothesis that cancer is the result of dysfunction from
a limited number of basic cellular processes common to all living human cells
was first introduced and later expanded on in a pair of classic essays by
Hanahan and Weinberg [@weinberg2000; @hallmarks_II]. In this view, all tumors
regardless of tissue of origin acquire six essential phenotypes; sustaining
proliferative signaling, evading growth suppressors, replicative immortality,
evading apoptosis, supportive angiogenesis, tissue invasion and metastasis
[@weinberg2000]. Advancements in the field over the ensuing decade led to the
inclusion of new tumor promoting processes, namely deregulation of cellular
energy cycle, tumor promoting inflammation, evasion of the immune system, genome
instability, and on top of these acquired traits a new appreciation for the
concept of the tumor microenvironment and its contribution to the evolution and
promotion of metastatic disease [@hallmarks_II].

@@

We find that most tumors segregate into one of twelve biologically distinct
clusters, regardless of the cancer type (**figure 1G,I** and **supplementary
figure S1**). Some cancer types ( _e.g._ pancreatic adenocarcinomas in the RAS
mutant clusters 8 and 9, and Kidney Chromophobe in cluster 1, see **supplementary
figure S1**) are unevenly distributed, though we could not identify any cancers
that were exclusive to a single tumor class. Only cancer types with the fewest
samples were found to be absent from one or more classes at all. Somewhat
surprisingly to us, this finding extends to histological subtypes of breast,
head and neck cancers, leukemias, _etc._ (see **Figure 1J** and **Supplemental
Figure S2**). This result implies that many, if not most, histological subtypes
could reflect differences in cell-of-origin, rather than fundamental differences
in proximal cause. The four major subtypes of breast cancer correspond to histological
and molecular expression profiles that define them and how they respond to experimental
stimuli [@breast-subtypes2]. It has been hypothesized that differences in the
molecular regulators of development in the precursor cell types present in the
breast epithelium drive the various histological phenotypes [@breast-subtypes;
@breast-subtypes3]. Consistent with this view we found that breast tumor samples
of the Luminal A subtype were heavily biased toward membership in clusters 1 and
7, and basal tumor samples were biased toward cluster 6 (**figure 1J**).
However, both subtypes also contained samples in every other class (without
exception), and Luminal B and Her2 positive samples are distributed across
clusters. We interpret these data to mean that inherited cell-of-origin
molecular profiles may predispose certain precursor cell types within the breast
epithelium to forming tumors of one class or another but are not determinative;
this view is compatible with the previously stated hypothesis but opens the way
for a more granular view of individual tumors. It would be surprising if we did
not observe bias for some cancers and subtypes amongst our classes, since some
treatment regimens have greater efficacy for patients
of a given cancer or histological subtype [@breast-subtypes2]. Nonetheless, the
basis for some tumors being treatment-refractory in spite of receiving the
standard of clinical care for diagnostic markers remains elusive. Doubtless some
of this is due to bad luck, as tumors can metastasize and remain dormant years
before they are detected at distal sites, or resistant clones may have already
arisen at undetectable levels [@hallmarks_II], but our analysis suggests the
possibility of identifying more informative molecular, histological or cellular
subtypes that could form a basis for future separation of patients into
different precision treatment regimens.

@@

Finally, we must remark on the limitations of our work exemplified in two of the
classes, 1 and 10, for which we did not find distinctive associations with
pathways (**figure 2B, C**). Class 1 has a relatively low proportion of mutated
pathways across the board, although it is broadly enriched in many of the same
tumor-promoting pathways common to the other groups. This likely does not
reflect a true difference in stages of tumor evolution, as our data show clearly
that this class is as likely to contain stage IV metastatic tumors as it is to
contain those of stage I (**figure 1L, M**). This class likely represents a
group of tumors with aberrations in methylation, copy number alterations,
structural variants, or combinations thereof which we excluded from our
analysis. Consistent with this, kidney chromophobe and thyroid cancers have high
proportions of SVs _vs._ other variant types [@PCAWG-MAIN] and are heavily
skewed to cluster 1 membership (**Supplementary figure S1**). Likewise, Class 8
represents a group of hyper-mutated tumors that harbor so many mutations that
virtually no pathway is unaffected. It seems likely that a significant fraction
of the samples flagged as “mutant” for each pathway are actually unaffected,
consistent with those samples have low information content. This could be
addressed with application of ever-increasingly sophisticated bioinformatics
algorithms for filtering likely passenger mutations [ _e.g._ @spliceAI;
@deep-learning-clinical-variants]. In the future, we hope to incorporate these
other data into a comprehensive pathway-centric analysis as we have done here
for point mutations and indels.

@@

Our findings imply that there are separate processes in the etiology of cancer
that can be broadly thought of as general cancer promoting, class-specific
mutations and metastasis. General cancer promoting processes must include
factors that relate to genome stability and immortality, as "enabling
characteristics" of the cancer phenotype [@hallmarks_II]. Such pathways are
disrupted in all or most of the defined classes (**supplementary figure S1**)
and often are the result of aberrations involving common driver genes such as
BRCA1/2, MMR genes, mitotic checkpoints, Cohesion complexes, _etc._
(**supplementary figure S3** and **supplemental file 1**). Class-specific
evolution must involve the acquisition of disruptions to pathways that may
individually be harmful ( _e.g._ highly proliferative cells are more likely to
senesce) but together produce more specialized cancer phenotype and increased
fitness (as in **figure 4** and **tables 1** and **ST2**). Importantly, our
observations do not imply the order in which these mutations should accumulate.
This could be addressed in a future study by integrating our analysis with
evolutionary analysis of clonality, drawing inferences from variant allele
frequencies as in @PCAWG-EVO. However, since many of the genes in the
non-cluster-specific pathways involve genome stability, it is reasonable to
hypothesize that these mutations promote or enable the acquisition of
class-specific defects via random mutation and natural selection, thus producing
the classes we observed. In support of this, the pan-cancer analysis of whole
genomes consortium (PCAWG) found that a subset of known oncogenic driver
mutations are highly enriched in early arising clones, whereas later arising
clones have much greater diversity in driver mutations [@PCAWG-EVO]. Moreover,
driver genes that are known to be responsible for discrete mutation signatures
such as APOBEC, BRCA1 and BRCA2 produce mutational hotspots reflecting the
varying selective pressures in different tissues [@PCAWG-MAIN].

One of the drawbacks of bulk tumor whole genome sequencing data is the problem
of tumor heterogeneity. Consortium samples are likely to contain contamination
from support tissue, stroma, inflammatory, and immune cells of the innate and
adaptive immune systems, all potentially harboring cancer supporting mutations
[@bhowmick_review]. We think it will be instructive to explore these ideas in
the context of single cell experiments.

@@

We report that pathway enrichment in metastatic tumors across all classes
(pooled) yielded generally lower effect sizes and larger p-values than the
cluster specific analysis, suggesting that signal is diluted when the samples
are analyzed together, and supporting the view that metastasis has
class-specific requirements. Since the number of metastatic samples is
relatively low this part of our analysis is likely underpowered and subject to
expanded analysis with larger cohorts. Nonetheless, the class-specific
metastatic pathway enrichment we observe (in **table 2**) could be a result of
the pathways directly promoting metastatic phenotypes in these classes, or
because they acquire a stronger or more complete oncogenic phenotype, as we
don’t know the order in which they were acquired. The fact that most metastatic
enrichment is cluster-specific and has a tendency to overlap with
cluster-specific pathways from non-metastatic tumors of neighboring clusters
favors the latter interpretation, and suggests a mechanism of complementarity,
in which newly acquired mutations result in similar classes converging on a
single deadly phenotype that captures most of the critical features of end-stage
cancer. One explanation for this could be that tumors must accumulate mutations
in a certain order or face elimination due to any of the selective bottlenecks
they face (senescence, the immune system, _etc._) [@hallmarks_II]. Some clusters
were characterized by cluster-specific pathways that seem superficially
unrelated; for example cluster 2 has mutations in the proteasome subunits,
cluster 3 has mutations impacting protein translation, whereas cluster 5 has
defects in RNA metabolism and clusters 7, 8 and 9 have defects in the more
traditional extra-cellular signaling pathways (as exemplified in **table 1**).
It may be that the primary effects of mutations in protein and RNA metabolism in
clusters 3 and 5 point to currently unappreciated mechanisms impacting hallmark
pathways. Clusters 3 and 5 are also marked by a decidedly different set of
pathways in metastasis. Clusters 3 and 5 notably have (in addition to signaling
pathways) mutations in glucagon and glycogen metabolism, respectively, in their
metastases (**table 2**), consistent with energy metabolism as one of the new
hallmarks [@hallmarks_II], but also distinguishing these classes from the others
in the order or nature of hallmark acquisition. Thus, even at relatively low
power, our analysis of metastasis further supports the pathway-centric approach
as a fruitful way to uncover real differences in underlying biology. As a major
caveat to this, we acknowledge that many therapies _do_ target highly-specific
driver genes, markers, and signaling pathways ( _e.g._ TP53, EGFR or HER2), but
understanding the broader context of the genetic background and pathway
vulnerability of tumors containing such markers may aid in creating smarter
combination therapies. We submit that when we discover the requirements of each
cluster with respect to pathway disruptions and metastasis we may be able to
target them therapeutically and prevent further adaptation.

### 2020-07-27 text/survival-methods.Rmd

We modeled baseline longevity in the U.S. with these different factors.
<mark>Key assumptions? (Room here for further descriptions and equations)</mark>
Using this base survival model, we constructed a likelihood function with
parameters describing the overall lethality of each cancer type as a modifier
$k_{tissue}$ of the rate parameter $r_{0}$, and cluster specific modifiers of
$k$ and optimized model fit using stan[@stanref]. The $k$ in our model can be
taken as, roughly, the death rate relative to the baseline rates as depicted in
**supplementary figure S5**. Thus, a value of $k=2$ means that a given cancer is
twice as deadly.

With an accurate model for baseline longevity in place, we sought to model
cancer as an event based modifier of long-term survival. Thus any factors or
events such as cancer would modify the baseline risk function as a multiplier.
The resulting coefficients can thus be interpreted as increasing baseline risk
by <mark>x-fold.</mark> Baseline risk was modeled as cumulative survival for
female vs. male caucasians, blacks and asians (hispanics are not clearly marked
in the tumor data; see **supplementary figure S5** for comparisons of raw CDC
data and baseline risk models). Thus we are able to accurately model baseline
risk of death taking these factors into account. The data themselves encompass
all other causes of death for these populations including cancer, thus freeing
our model from concerns about other confounding factors ( _e.g._ Diabetes). For
demonstration purposes we also illustrate some examples of the expected survival
for a population that receives a diagnosis with doubled risk of death at ages
40, 50, 60, 70, and 80 (**supplementary figure S6**).

As an aside, we did not explicitly account for stage at diagnosis in our model
for the following reasons. In the tumor sample data, stage at diagnosis is
confounded with cancer type because some cancers are screened agressively (
_e.g._ colorectal and prostate cancers) while others are diagnosed typically
after they become problematic for the patient's lifestyle ( _e.g._ ovarian and
pancreatic cancers). Secondly, such a model specification would suffer from
added noise because the staging data are not well standardized across cancer
types, have different criteria, and because it is unclear what the relationship
between stage and advancement of disease is (for example some sub-stage 4 tumors
are metastatic). To compound this latter issue, our tumor set is vastly
under-powered given the uneven representation of stage across cancer types.

>>>>

In order to understand how the diagnosis of cancer changes the expectation for
longevity of patients, we first began by modeling the longevity of the overall
population of the US based on CDC life tables. Although childhood cancers exist
and are tragic, most cancer patients have diagnosis later in life. We modeled
baseline longevity using a risk rate function:

$$R(t) = R_{20}\exp(a(A-20) + D(t) k a (\frac{t}{365}))$$

Time is split into two components. $A$ going from 0 up to the age of the patient
at diagnosis and then remaining fixed, whereas $t$ starts at 0 upon diagnosis
and represents the days since diagnosis, a field recorded in the dataset. $D(t)$
is an indicator for whether diagnosis has occurred or not (before diagnosis D(t)
= 0, after it equals 1). The model essentially assumes that the risk of dying
per year at age 20 years is $R_{20}$ and that it increases exponentially with a
constant rate $a$ up to diagnosis, and then with a new rate $ka$
thereafter. Though the model does extend back to earlier ages, relatively few
patients were under age 20.

Given this risk per unit time, the probability to die at time $T+dT$ is the
probability to survive to time $T$ which is $(1-P(T))$ and then die in the
remaining interval which is $R(T)dT$

$$dP = (1-P(T)) R(T) dT$$

leading to the differential equation for the cumulative probability of death
$P(T)$

$$\frac{dP}{dT} = R(T)(1-P(T))$$

The solution of this ODE is:

$$P(T) = 1 - \exp(-\int_0^T R(t) dt)$$

And the density of deaths per unit time is the derivative with respect to T:

$$p(T) = R(T) \exp(-\int_0^T R(t)dt)$$

The $R(T)$ function has two important parameters, the $a$ value which represents
the background risk rates, and the $k$ value, which is a function of both cancer
tissue type, and the cancer genomic class.

$$k = k_{tis} k_{cl}$$

The product $k_{tis}k_{cl}$ is obviously symmetric between the two factors. To
disambiguate the meaning of the two, the prior distribution of $k_{cl}$ has peak
probability at 1 and a relatively narrow width, due to the fact that it is a
tissue specific multiplicative modifying factor, this allows the overall
magnitude of k to be primarily determined by the $k_{tis}$ value whose prior is
significantly less constrained so that there is a wider range of dangers across
tissue types.

The $k$ in our model can be taken as, roughly, the aging rate relative to the
baseline rates as depicted in **supplementary figure S5**. Thus, a value of
$k=2$ means that a given cancer causes you to experience the same risk in time
$dt$ as a non-cancer patient would experience at the same age, in time $2dt$.

Using this model, we split the data by Male/Female sex, and by White, vs Black
race as these were the main well-identified categories available to us in the
dataset. Sex is known to be a risk factor for death, with males dying at
slightly higher rates for all ages of interest in our dataset. Black vs White
race is an important category due to differences in access to care and other
socioeconomic and education related factors.

Using the model described above we fit the data using Stan via the rstan package
in R. Posterior inferences are shown in **FIXME WHICH FIGURE?**.

**supplementary figure S5** gives a comparison on log-scale between the post-fit
survival curves and the CDC data it is based on. For demonstration purposes we
also illustrate some examples of the expected survival for a population that
receives a diagnosis with doubled risk of death at ages 40, 50, 60, 70, and 80
(**supplementary figure S6**).

As an aside, we did not explicitly account for stage at diagnosis in our model
for the following reasons. In the tumor sample data, stage at diagnosis is
confounded with cancer type because some cancers are screened aggressively (
_e.g._ colorectal and prostate cancers) while others are diagnosed typically
after they become problematic for the patient's lifestyle ( _e.g._ ovarian and
pancreatic cancers). Secondly, such a model specification would suffer from
added noise because the staging data are not well standardized across cancer
types, have different criteria, and because it is unclear what the relationship
between stage and advancement of disease is (for example some sub-stage 4 tumors
are metastatic). To compound this latter issue, our tumor set is vastly
under-powered given the uneven representation of stage across cancer types.
